Novel Methods for Characterizing Peaks from Direct Injection FT-MS Experiments

### 2019-07-26 peakcharacterization_manuscript.Rmd
updating some of the language

The data consists of profile mode M/Z spectra from multiple scans encoded as M/Z and intensity values for each scan. There is no information about the original frequency values available in either the `raw` files or the `mzML` files. However, the frequency values can be approximated by averaging the M/Z of adjacent points, and dividing the M/Z by the difference. A constant difference of adjacent frequency values is expected, and points that have frequency differences within a narrow range can be used to construct a linear model relating M/Z to frequency for individual scans based on the formula:

$$frequency = intercept + slope* \frac{1}{\sqrt{mz}} $$
A model is generated for each scan, and then a single model using the median values of the intercept and slope across scans is chosen, and applied to convert the M/Z values to frequency for all scans. A single model rather than scan specific models is used because this was found to maintain the relative ordering of the peaks in M/Z and frequency space. Integer values (for use with the IRanges R package) are generated by multiplying each frequency by 400 and rounding.

@@

To enable interval arithmetic, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package. The sliding and tiled windows are also converted to IRanges objects the same way.

@@

Within each initial interval region, peaks in each scan are detected (see **Peak Finding**), and their centers binned by the tiled windows. Adjacent tiled windos with non-zero peak counts are merged together, and any zero peak count tiled windows split the initial region into multiple peak interval regions. These interval regions should contain a single **real** peak that was detected in one or more scans.

@@

On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package. These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either M/Z or frequency), and where the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.

>>>>

The data consists of profile mode M/Z spectra from multiple scans encoded as M/Z and intensity values for each scan. There is no information about the original frequency values available in either the `raw` files or the `mzML` files. However, the frequency values can be approximated by averaging the M/Z of adjacent points, and dividing the M/Z by the difference. The *true* difference can be obtained by examining the mode of the mean frequency differences, and then constraining *useful* points to be within 2% of the mode value. These points can be used to construct a linear model relating M/Z to frequency for individual scans based on the formula:

$$frequency = intercept + slope* \frac{1}{\sqrt{mz}} $$
A model is generated for each scan, and then a single model using the median values of the intercept and slope across scans is chosen, and applied to convert the M/Z values to frequency for all scans. A single model rather than scan specific models is used because this was found to maintain the relative ordering of the peaks in M/Z and frequency space. Integer values (for use with the IRanges R package) are generated by multiplying each frequency value by 400 and rounding.

@@

To enable interval arithmetic, the frequency points are converted to single width intervals by multiplying and rounding (a multiplier of 400 in this work) and storing them as IRanges objects from the IRanges Bioconductor package [citation]. The sliding and tiled windows are also converted to IRanges objects the same way.

@@

Within each initial interval region, peaks in each scan are detected (see **Peak Finding**), and their centers binned by the tiled windows. Adjacent tiled windows with non-zero peak counts are merged together, and any zero peak count tiled windows split the initial region into multiple peak interval regions. These interval regions should contain a single **real** peak that was detected in one or more scans.

@@

On a single scan level, possible peaks are detected by simple bump-hunting for two increasing points followed by two decreasing points using the `find_peaks` function in the pracma package [citation]. These possible peaks are then characterized using a weighted parabolic fit of log-intensity to position (where position is either M/Z or frequency), and where the weights for each point are the relative log-intensity compared to the maximum log-intensity for the peak.
